\section{Grothendieck Categories}

\subsection{Definition}

A \textbf{Grothendieck category} $\mathcal{A}$ is an category satisfying the following conditions:

\begin{enumerate}
    \item $\mathcal{A}$ is an \textbf{abelian category}.

    \item $\mathcal{A}$ has \textbf{arbitrary (possibly infinite) coproducts}, i.e. for every family of objects $(X_i)_{i \in I}$, the coproduct
    \[
    \bigoplus_{i \in I} X_i
    \]
    exists in $\mathcal{A}$.

    \item $\mathcal{A}$ satisfies \textbf{AB5}:

    filtered colimits (direct limits) of short exact sequences are exact. That is, given a directed system of short exact sequences
    \[
    0 \to A_i \to B_i \to C_i \to 0,
    \]
    the induced sequence
    \[
    0 \to \varinjlim A_i \to \varinjlim B_i \to \varinjlim C_i \to 0
    \]
    is also exact.

    \item $\mathcal{A}$ has a \textbf{generator}. That is, there exists an object $G \in \mathcal{A}$ such that the functor
    \[
    \operatorname{Hom}(G,-): \mathcal{A} \to \mathbf{Set}
    \]
    is faithful.
\end{enumerate}

\subsection{Generator Condition (Equivalent Form)}

An object $G \in \mathcal{A}$ is a generator if and only if for every object $X \in \mathcal{A}$, there exists a set $I$ and an epimorphism
\[
G^{(I)} \twoheadrightarrow X,
\]
where
\[
G^{(I)} = \bigoplus_{i \in I} G
\]
is the coproduct of copies of $G$, indexed by the set $I$.

\subsection{Remarks}

\begin{itemize}
    \item The AB5 condition implies that filtered colimits behave well with respect to exact sequences.

    \item Coproducts in $\mathcal{A}$ generalize direct sums in module categories.

    \item The existence of a generator ensures that every object in the category can be ``built'' from a single object via coproducts and quotients.
\end{itemize}

\section{Basic Definitions Used in Grothendieck Categories}

\subsection{Abelian Categories}

A category $\mathcal{C}$ is called \textbf{abelian} if:

\begin{enumerate}
    \item $\mathcal{C}$ is \textbf{preadditive}, i.e. each hom-set $\operatorname{Hom}(A,B)$ is an abelian group and composition is bilinear.

    \item $\mathcal{C}$ has a \textbf{zero object}, denoted $0$, which is both initial and terminal.

    \item $\mathcal{C}$ has all \textbf{binary biproducts} (direct sums) $A \oplus B$.

    \item $\mathcal{C}$ has all \textbf{kernels} and \textbf{cokernels}.

    \item Every monomorphism and epimorphism is \textbf{normal}, i.e.
    \begin{itemize}
        \item every monomorphism is a kernel of some morphism,
        \item every epimorphism is a cokernel of some morphism.
    \end{itemize}
\end{enumerate}

\subsection{Coproducts}

Let $\mathcal{C}$ be a category and let $X_1, X_2$ be objects of $\mathcal{C}$.

A \textbf{coproduct} of $X_1$ and $X_2$ is an object
\[
X_1 \sqcup X_2
\]
together with morphisms
\[
i_1: X_1 \to X_1 \sqcup X_2,
\qquad
i_2: X_2 \to X_1 \sqcup X_2,
\]
satisfying the following universal property:

For every object $Y$ and morphisms
\[
f_1: X_1 \to Y,
\qquad
f_2: X_2 \to Y,
\]
there exists a unique morphism
\[
f: X_1 \sqcup X_2 \to Y
\]
such that the diagram commutes, i.e.
\[
f_1 = f \circ i_1,
\qquad
f_2 = f \circ i_2.
\]

\subsection{Universal Property Diagram}

\[
\begin{tikzcd}
X_1 \arrow[r, "i_1"] \arrow[dr, "f_1"'] & X_1 \sqcup X_2 \arrow[d, "f"] & X_2 \arrow[l, "i_2"'] \arrow[dl, "f_2"] \\
& Y &
\end{tikzcd}
\]

\subsection{Notation}

The coproduct of $X_1$ and $X_2$ is often written as:
\[
X_1 \sqcup X_2
\quad \text{or} \quad
X_1 \oplus X_2
\quad \text{or} \quad
X_1 + X_2.
\]

In abelian categories, coproducts and products coincide for finite families, and are called \textbf{biproducts}.

\subsection{Direct limits of short exact sequence}

Let $\{0 \to A_i \to B_i \to C_i \to 0\}_{i \in I}$ be a directed system of short exact sequences in an abelian category $\mathcal{A}$, indexed by a directed set $I$.
Taking direct limits (colimits) gives the induced sequence
\[
0 \longrightarrow \varinjlim_{i \in I} A_i
\longrightarrow \varinjlim_{i \in I} B_i
\longrightarrow \varinjlim_{i \in I} C_i
\longrightarrow 0.
\]
In a Grothendieck category (AB5), this sequence is again short exact.
Let $I$ be a directed set and let
\[
\left\{0 \to A_i \xrightarrow{f_i} B_i \xrightarrow{g_i} C_i \to 0 \right\}_{i \in I}
\]
be a directed system of short exact sequences in an abelian category $\mathcal{A}$.

For each $i \le j$ in $I$, assume we have compatible morphisms
\[
\alpha_{ij}: A_i \to A_j, \quad
\beta_{ij}: B_i \to B_j, \quad
\gamma_{ij}: C_i \to C_j
\]
such that all structure maps commute with the short exact sequences, i.e.
\[
\beta_{ij} \circ f_i = f_j \circ \alpha_{ij},
\qquad
\gamma_{ij} \circ g_i = g_j \circ \beta_{ij}.
\]

\subsection*{Step 1: Form the direct limits}

Take direct limits in $\mathcal{A}$:
\[
A := \varinjlim_{i \in I} A_i,
\qquad
B := \varinjlim_{i \in I} B_i,
\qquad
C := \varinjlim_{i \in I} C_i.
\]

Let the canonical maps be
\[
\alpha_i: A_i \to A,
\qquad
\beta_i: B_i \to B,
\qquad
\gamma_i: C_i \to C.
\]

\subsection*{Step 2: Induced morphisms}

By the universal property of colimits, the maps $f_i$ induce a unique morphism
\[
f := \varinjlim f_i : A \to B
\]
such that
\[
f \circ \alpha_i = \beta_i \circ f_i \quad \text{for all } i \in I.
\]

Similarly, the maps $g_i$ induce
\[
g := \varinjlim g_i : B \to C
\]
such that
\[
g \circ \beta_i = \gamma_i \circ g_i.
\]

\subsection*{Step 3: Resulting sequence}

We obtain an induced sequence
\[
A \xrightarrow{f} B \xrightarrow{g} C.
\]

\subsection*{Step 4: Exactness after taking limits}

In general abelian categories, the sequence need not be exact.

However, if $\mathcal{A}$ is a Grothendieck category (i.e. satisfies AB5), then filtered colimits are exact, and we obtain a short exact sequence:
\[
0 \to \varinjlim A_i \xrightarrow{\;\varinjlim f_i\;} \varinjlim B_i \xrightarrow{\;\varinjlim g_i\;} \varinjlim C_i \to 0.
\]

\subsection*{Summary}

The process is:
\[
\text{direct system of SES}
\;\Longrightarrow\;
\text{take colimits objectwise}
\;\Longrightarrow\;
\text{induced sequence}
\;\Longrightarrow\;
\text{short exact sequence (in AB5 categories)}.
\]

\section{Examples of Grothendieck Categories}

\subsection{Abelian Groups}

The category $\mathbf{Ab}$ of abelian groups is a Grothendieck category.

\begin{itemize}
    \item It is an abelian category.
    \item It has arbitrary coproducts (direct sums of abelian groups).
    \item It satisfies AB5: filtered colimits of short exact sequences of abelian groups are exact.
    \item It has a generator: the group $\mathbb{Z}$.
\end{itemize}

Indeed, for any abelian group $A$, there exists a set $I$ and a surjection
\[
\mathbb{Z}^{(I)} \twoheadrightarrow A,
\]
so $\mathbb{Z}$ is a generator.

\subsection{Modules over a Ring}

Let $R$ be an associative ring with identity (not necessarily commutative). The category
\[
\operatorname{Mod}(R)
\]
of right (or left) $R$-modules is a Grothendieck category.

\begin{itemize}
    \item It is abelian.
    \item It has arbitrary coproducts (direct sums of modules).
    \item It satisfies AB5.
    \item It has a generator: the free module $R$.
\end{itemize}

For every module $M$, there exists a set $I$ and an epimorphism
\[
R^{(I)} \twoheadrightarrow M.
\]

\subsection{Sheaves of Abelian Groups}

Let $X$ be a topological space. The category
\[
\mathbf{Sh}(X)
\]
of sheaves of abelian groups on $X$ is a Grothendieck category.

More generally, for any ring $R$, the category of sheaves of $R$-modules on $X$ is also Grothendieck.

\begin{itemize}
    \item It is abelian.
    \item It has arbitrary coproducts (computed sectionwise).
    \item It satisfies AB5 (filtered colimits are computed stalkwise and preserve exactness).
    \item It has a generator (constructed from skyscraper sheaves).
\end{itemize}

\subsection{Sheaves on a Ringed Space}

Let $(X, \mathcal{O}_X)$ be a ringed space. Then the category
\[
\mathcal{O}_X\text{-Mod}
\]
of sheaves of $\mathcal{O}_X$-modules is a Grothendieck category.

\begin{itemize}
    \item Abelian structure comes from sheaf morphisms.
    \item Coproducts and limits are computed sectionwise.
    \item AB5 holds via stalkwise exactness.
    \item A generator exists via direct sums of structure sheaves over open sets.
\end{itemize}

\subsection{Quasi-Coherent Sheaves}

Let $V$ be an algebraic variety, scheme, or algebraic stack. The category
\[
\operatorname{Qcoh}(V)
\]
of quasi-coherent sheaves is a Grothendieck category.

\begin{itemize}
    \item Abelian structure depends on the geometry of $V$.
    \item Arbitrary coproducts exist.
    \item Filtered colimits are exact.
    \item A generator exists locally via affine charts.
\end{itemize}

\subsection{Sheaves on a Site}

Let $(\mathcal{C}, J)$ be a small site, i.e. a small category $\mathcal{C}$ equipped with a Grothendieck topology $J$.

Then the category
\[
\mathbf{Sh}(\mathcal{C}, J)
\]

of sheaves of abelian groups on the site is a Grothendieck category.

\begin{itemize}
    \item It is abelian.
    \item It has all coproducts.
    \item It satisfies AB5.
    \item It has a generator given by representable sheaves (via Yoneda embedding and sheafification).
\end{itemize}